---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidyassessments

<!-- badges: start -->
[![R-CMD-check](https://github.com/barnabywalker/tidyassessments/workflows/R-CMD-check/badge.svg)](https://github.com/barnabywalker/tidyassessments/actions)
<!-- badges: end -->

This is a package to make some models used for occurrence-based automated extinction risk assessments work with the tidymodels workflow.

This package was developed to help the analysis for a paper developing [guidelines for automated conservation assessments](https://github.com/barnabywalker/guidelines-for-automated-assessments). To add new models to tidymodels, it's best to package them up so they work in parallel across platforms.This is a package to make some models used for occurrence-based automated extinction risk assessments work with the tidymodels workflow.

This package was developed to help the analysis for a paper developing [guidelines for automated conservation assessments](https://github.com/barnabywalker/guidelines-for-automated-assessments).

So far, the models implemented produce binary classifications of species as "threatened" or "not threatened".

* Rule-based models, using fixed thresholds on computed metrics:
  * IUCN threshold - a single threshold on the Extent of Occurrence (EOO) of a species.
  * ConR - thresholds on EOO, area of occupancy (AOO), and number of locations intended to
    approximate an IUCN Red List criterion B assessment.

## Installation

You can install the development version of tidyassessments like so:

``` r
remotes::install_github("barnabywalker/tidyassessments")
```

## Example

You can set up a tidymodels model specification like this:

```{r example}
library(tidyassessments)

spec <- rule_based(engine="eoo")
```

